Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-08 paper/paper2.tex
abbrevs.
Typo fixes and added a single new citation

\section{Theory}
\subsection{Markov State Models}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use. Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g. 4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2011Markov, Bowman2012Equilibrium, Beauchamp2012Simple}. Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established. The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{someone}.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set.
\begin{equation}
  \label{eq:bic}
  \mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
  \label{eq:aic}
  \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.  Model selection is performed by a minimization of the criterion. The BIC is derived based on the dominant terms in the Laplace approximation to the logarithim of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}. The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series. Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.

@@

\begin{algorithm}
\caption{Monte Carlo Estimation of the State Volumes}
\label{alg:volumes}
\begin{algorithmic}[1]
\Procedure{MC Volume Estimation}{$\sigma, A, M$}
	\State $\mathbf{c} \gets [0, 0, \ldots, 0]$\Comment{$\mathbf{c}$ is a vector of length $k$}
	\While{$\sum_{i} c_i \leq M$}
		\State $\mathbf{x} \gets $ sample from axis-aligned hyper cube containing $A$
		\If{$\mathbf{x} \in A$}
			\State $c_{\sigma(\mathbf{x})} \gets c_{\sigma(\mathbf{x})} + 1$
		\EndIf
	\EndWhile
	\State $\mathbf{v} \gets \mathbf{c} \cdot(\sum_{i=1}^k c_i)^{-1}$ \Comment{$\mathbf{v}$ contains the relative volumes of each state}
	\State \textbf{return} $\mathbf{v}$
\EndProcedure
\end{algorithmic}
\end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
\subsection{Simulation Protocol}

@@

Two systems were investigated using this likelihood scheme. The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
$$
\frac{d\mathbf{x}}{dt} = - \nabla V(\mathbf{x})\zeta + \sqrt{2kT\zeta} R(t),
$$ where $\zeta = 10^{-3}$, $kT = 15$, $R(t)$ is a delta-correlated Gaussian process with zero mean, and $V(\mathbf{x})$ was defined as
$$
V(\mathbf{x}) =  \sum_{j=1}^4 A_j \cdot \exp\Big(a_j (x_1-X_j)^2+b_j(x_1-X_j)(x_2-Y_j)+c_j(x_2-Y_j)^2\Big),
$$ where $a = (-1, -1, -6.5, 0.7)$; $b = (0, 0, 11, 0.6)$; $c = (-10, -10, -6.5, 0.7)$; $A = (-200, -100, -170, 15)$; $X = (1, 0, -0.5, -1)$; $Y = (0, 0.5, 1.5, 1)$ as suggested by \citet{Muller1979Location}. Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps.

@@

Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{Lindorff2010Improved} TIP3P water model\cite{Jorgensen1983Comparison}). We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory. To calculate the tICs, each conformation was represented as a vector of distances between all pairs of residues separated by at least three amino acids. The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows. Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$. Then we define
$$
d(A,B) \equiv \min_{i, j} ||a_i - b_j||_2,
$$ where $|| \cdot ||_2$ denotes the $\ell_2$ norm. The tICs were computed using a correlation lag time of 200 ns. We also performed Principal Components Analysis (PCA) on the same residue-residue distance representation and built MSMs using the top three PCs. In each projection, $k$-centers was used with the Euclidean distance metric.

@@

Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system. For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the dataset during the fitting of an MSM. For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting. Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.

@@

As was the case with the M\"{u}ller MSMs, the log-likihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}). The test log likelihood has a maximum at 650 states, and decreases with the addition of further states. This trend is matched by the AIC, which is computed only on the training data, while the BIC on the other hand seems to over-penalize model complexity. The implied timescales of these MSMs converge rapidly, and selection based on their convergence is consistent with the statistical methods.

>>>>

\section{Theory}
\subsection{Markov State Models}
Let $\mathbf{x}_t$ be a ergodic time-homogeneous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.

@@

Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use. Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g. 4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2011Markov, Bowman2012Equilibrium, Beauchamp2012Simple}. Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established. The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{bowman2009progress}.

@@

Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}. Cross-validation attempts to directly measure the Markov model's generalization error. First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion. This scheme can be repeated many times with different partitions of the data set.
\begin{equation}
  \label{eq:bic}
  \mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
\end{equation}
\begin{equation}
  \label{eq:aic}
  \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.  Model selection is performed by a minimization of the criterion. The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}. The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series. Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.

@@

\begin{algorithm}
\caption{Monte Carlo Estimation of the State Volumes}
\label{alg:volumes}
\begin{algorithmic}[1]
\Procedure{MC Volume Estimation}{$\sigma, A, M$}
	\State $\mathbf{c} \gets [0, 0, \ldots, 0]$\Comment{$\mathbf{c}$ is a vector of length $k$}
	\While{$\sum_{i} c_i \leq M$}
		\State $\mathbf{x} \gets $ sample from axis-aligned hyper cube containing $A$
		\If{$\mathbf{x} \in A$}
			\State $c_{\sigma(\mathbf{x})} \gets c_{\sigma(\mathbf{x})} + 1$
		\EndIf
	\EndWhile
	\State $\mathbf{v} \gets \mathbf{c} \cdot(\sum_{i=1}^k c_i)^{-1}$ \Comment{$\mathbf{v}$ contains the relative volumes of each state}
	\State \textbf{return} $\mathbf{v}$
\EndProcedure
\end{algorithmic}
\end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
\subsection{Simulation Protocol and MSM Construction}

@@

Two systems were investigated using this likelihood scheme. The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
$$
\frac{d\mathbf{x}}{dt} = - \nabla V(\mathbf{x})\zeta + \sqrt{2kT\zeta} R(t),
$$ where $\zeta = 10^{-3}$, $kT = 15$, $R(t)$ is a delta-correlated Gaussian process with zero mean, and $V(\mathbf{x})$ was defined as
$$
V(\mathbf{x}) =  \sum_{j=1}^4 A_j \cdot \exp\Big(a_j (x_1-X_j)^2+b_j(x_1-X_j)(x_2-Y_j)+c_j(x_2-Y_j)^2\Big),
$$ where $a = (-1, -1, -6.5, 0.7)$; $b = (0, 0, 11, 0.6)$; $c = (-10, -10, -6.5, 0.7)$; $A = (-200, -100, -170, 15)$; $X = (1, 0, -0.5, -1)$; $Y = (0, 0.5, 1.5, 1)$ as suggested by \citet{Muller1979Location}. Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps. The initial positions were sampled via a uniform distribution over the box: $[-1.5, 2.0] \times [-0.2, 2.0]$.

@@

Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E. Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{Lindorff2010Improved} TIP3P water model\cite{Jorgensen1983Comparison}). We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory. To calculate the tICs, each conformation was represented as a vector of distances between all pairs of residues separated by at least three amino acids. The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows. Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the Cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$. Then we define
$$
d(A,B) \equiv \min_{i, j} ||a_i - b_j||_2,
$$ where $|| \cdot ||_2$ denotes the $\ell_2$ norm. The tICs were computed using a correlation lag time of 200 ns. We also performed Principal Components Analysis (PCA) on the same residue-residue distance representation and built MSMs using the top three PCs. In each projection, $k$-centers was used with the Euclidean distance metric.

@@

Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system. For large biomolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the data set during the fitting of an MSM. For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting. Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.

@@

As was the case with the M\"{u}ller MSMs, the log-likelihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}). The test log likelihood has a maximum at 650 states, and decreases with the addition of further states. This trend is matched by the AIC, which is computed only on the training data, while the BIC on the other hand seems to over-penalize model complexity. The implied timescales of these MSMs converge rapidly, and selection based on their convergence is consistent with the statistical methods.
